---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
require(rmarkdown)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EWSmethods

<!-- badges: start -->
<!-- badges: end -->

*EWSmethods* is a user friendly interface to various methods of performing Early Warning Signal (EWS) assessments. This R package allows the user to input univariate or multivariate data and perform either traditional rolling window or expanding window EWS approaches. Publication standard and ggplot inspired figures can also be generated during this process. *EWSmethods* also provides an R interface to [__EWSNET__](https://ewsnet.github.io), a deep learning modelling framework for predicting critical transitions.

## Installation

You can install the development version of EWSmethods from [GitHub](https://github.com/duncanobrien/EWSmethods) with:

``` {r}
# install.packages("devtools")
#devtools::install_github("duncanobrien/EWSmethods")
```

## Examples

Imagine we have 50 years of monitoring data for a local population of skylarks (*Alauda arvensis*) as well as mean body mass data. We could calculate either rolling or expanding window EWSs for the abundance data as would do so using EWSmethods as below:

```{r roll_example, fig.show='hide'}
library(EWSmethods)
set.seed(125) #seed to ensure reproducible results

skylark_data <- data.frame(time = seq(1:50), abundance = rnorm(50,mean = 100,sd=20), trait = rnorm(50,mean=40,sd=5)) #dummy skylark dataset

ews_metrics <- c("SD","ar1","skew") #the early warning signal metrics we wish to compute

roll_ews <- univariate_EWS_wrapper(data = skylark_data[,1:2],metrics =  ews_metrics,method = "rolling",winsize = 50, ggplotIt =TRUE, y_lab = "Skylark abundance") #lets using a rolling window approach

print(roll_ews$EWS$cor) #return the Kendall Tau correlations for each EWS metric
```

We can then use the resulting figures to identify oncoming transitions. In this case, we  expect no transition as the data is randomly sampled from a normal distribution and this is evident in the Kendall Tau values, with no strong positive correlation with time: 

```{r rolling_plot, dpi=300, echo = FALSE}
roll_ews$plot
```
Alternatively, we may be more interested in expanding windows as that approach standardises the changing EWS metrics over time and therefore allows the strength of multiple signals to be combined. We could achieve this using the following code:

```{r _expand_example, fig.show='hide'}
exp_ews <- univariate_EWS_wrapper(data = skylark_data[,1:2],metrics =  ews_metrics,method = "expanding",burn_in = 10, threshold = 2, tail.direction = "one.tailed", ggplotIt =TRUE, y_lab = "Skylark abundance") #lets use a rolling window approach

head(exp_ews$EWS) #return the head of the EWS dataframe
```

And again, we can then use the resulting figures to identify oncoming transitions. Whilst we have some trangressions of the 2$\sigma threshold, we only consider these signals "warnings" if two or more consecutive signals are identified (Clements *et al.* 2019).

```{r expanding_plot, dpi=300, echo = FALSE}
exp_ews$plot
```
A second benefit of the expanding window approach is that additional information can be used to improve the reliability of the assessment. For example, in our hypothetical skylark dataset, we have measured average population body mass. This data can then be delivered to *EWSmethods* using the `trait` argument.

```{r expand_trait_example, fig.show='hide'}

trait_metrics <- c("SD", "ar1", "trait")
exp_ews_trait <- univariate_EWS_wrapper(data = skylark_data[,1:2],metrics =  trait_metrics,method = "expanding",burn_in = 10, threshold = 2, tail.direction = "one.tailed", ggplotIt =TRUE, y_lab = "Skylark abundance",
trait = skylark_data$trait, trait_lab = "Body mass (g)", trait_scale = 5)

```
```{r expanding_plot_trait, dpi=300, echo = FALSE}
exp_ews_trait$plot
```


You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.
