---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EWSmethods

<!-- badges: start -->
<!-- badges: end -->

*EWSmethods* is a user friendly interface to various methods of performing Early Warning Signal (EWS) assessments. This R package allows the user to input univariate or multivariate data and perform either traditional rolling window or expanding window EWS approaches. Publication standard and ggplot inspired figures can also be generated during this process. *EWSmethods* also provides an R interface to __EWSNET__ ([link](https://ewsnet.github.io)), a deep learning modelling framework for predicting critical transitions.

## Installation

You can install the development version of EWSmethods from [GitHub](https://github.com/duncanobrien/EWSmethods) with:

``` {r}
# install.packages("devtools")
#devtools::install_github("duncanobrien/EWSmethods")
```

## Examples

Imagine we have 50 years of monitoring data for a local population of skylarks (*Alauda arvensis*) as well as mean body mass data. We could calculate either rolling or expanding window EWSs for the abundance data as would do so using EWSmethods as below:

```{r example}
library(EWSmethods)

skylark_data <- data.frame(time = seq(1:50), abundance = rnorm(50,mean = 20), trait = rnorm(50,mean=40,sd=5)) #dummy skylark dataset

ews_metrics <- c("SD","ar1","skew") #the early warning signal metrics we wish to compute

roll_ews <- univariate_EWS_wrapper(data = skylark_data[,1:2],metrics =  ews_metrics,method = "rolling",winsize = 50, ggplotIt =TRUE) #lets using a rolling window approach

print(roll_ews$EWS$cor) #return the Kendall Tau correlations for each EWS metric

```

```{r rolling_plot, echo = FALSE}
roll_ews$plot
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
