---
title: "Alternative Resilience Measures"
author: "Duncan O'Brien"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alternative Resilience Measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About this tutorial

This tutorial introduces resilience metrics distinct from critical slowing down based/Early Warning Signal indicators. The primary benefit of this collection of techniques is that they are both model and equilibrium free; they don't assume a system is a equilibrium.

`EWSmethods` provides three distinct approaches to quantifying system resilience. The tutorial will go through each in turn and give examples of their usage.

Greater detail on each function can be found at the [Reference](https://duncanobrien.github.io/EWSmethods/reference/index.html) page.

<br>  

# Getting started

```{r, message=FALSE}
set.seed(123) #to ensure reproducible data

library(EWSmethods)
library(ggplot2) #for plotting
```

We will use the same multi-species community described in the [Performing Early Warning Signal Assessments](https://duncanobrien.github.io/EWSmethods/articles/ews_assessments.html) tutorial.

Briefly, the data object `"simTransComms"` contains three replicate datasets of a simulated five species community that has been driven to transition by the introduction of an invasive species (following Dakos 2018). The strength of impact by the invasive species gradually increases through time between *t100* and *t200* at which point it is held constant.

Lets load the data in:
```{r load_data, eval=T}
data("simTransComms")
```

and plot one of the communities. The vertical dashed line represents a sudden transition due to the invasive species.

```{r vis_data,dpi=144}
matplot(simTransComms$community1[,3:7], type = "l", xlab = "Time", ylab = "Density",  main = "Transitioning five species community")
abline(v=simTransComms$community1$inflection_pt,col = "grey60", lty = "dashed")
```

Now we have a system/community that is experiencing some form of stress and we can apply univariate or multivariate resilience measures to give us an idea of how the system is responding to that stress.

<br>  

# 1) Index of Variability

The simplest measure provided by `EWSmethods` is the multivariate index of variability (MVI - Brock and Carpenter, 2006). The MVI is represented by the square root of the dominant eigenvalue of the covariance matrix of all species.

<br>  

### _Interpretation_ 
Increasing MVI represents decreasing stability.

<br>  

### _Usage_ 
Provide the community data (first column being time, the remainder being species values) to the `mvi()` function, and specify a `winsize` to calculate the index over. The default `winsize` is 50\% the total length of the time series, but we have set `winsize = 25` to ensure that the periods before and after the sudden change in \code{simTransComms$community1} are distinct. 

```{r mvi}
egMVI <- mvi(data = simTransComms$community1[,2:7], winsize = 25)
```

If we then plot the index, we can see that before and after the 'inflection point' (where the community goes through a transition) there are clear differences in stability. 

What is also apparent is that the MVI begins to increase prior to the sudden change, albeit only slightly. 

```{r mvi2,dpi=144}
mvi_plot_data <- merge(simTransComms$community1,as.data.frame(egMVI),by="time") # combine the mvi data with the raw time series

ggplot(mvi_plot_data,aes(x = time, y = mvi)) + geom_line() + geom_vline(xintercept = mvi_plot_data$inflection_pt,col = "grey60", linetype = "dashed") + theme_bw() + xlab("Time") + ylab("MVI value")
```

<br>  

### _Caveats_ 
The MVI only quantifies the variability of the community rather than changing relationships between community components. It is therefore susceptible to data quality, the inherent stochasticity of the system, and rate of forcing.

<br>  

# 2) S-map Jacobians

The cutting edge of resilience metrics belong to approaches which exploit empirical dynamic modelling (EDM - Sugihara *et al.* 2012). EDM exploits Takens’ embedding theorem (Takens 1981) which suggests that an underlying latent system/attractor manifold can be reconstructed from one or more related time series. By lag embedding two or more time series, their interaction strength (or causal strength) can be estimated. 

This approach consequently requires no knowledge of the underlying equations nor numerical estimates of the system's equilibrium state as the local Jacobians are directly estimated from the time series. The local Lyapunov stability is calculated as the dominant eigenvalue of the estimated Jacobian matrix.

Jacobians consist of the first-order partial derivatives for a multivariate function and are critical to resilience quantification. They are so important because they allow us to differentiate two or more functions with respect to two or more variables, something not achievable using standard derivation. Jacobian matrices therefore provide insight in to how each variable (or species) in a system's master equation will change in the successive time step; if we know the true Jacobian, we know the future.

`EWSmethods` provides access to two forms of s-map estimated Jacobians, one for univariate data (following Grziwotz *et al.* 2023) and another for multivariate data (following Ushio *et al.* 2018).

<br>  

### _Interpretation_
A local Lyapunov stability value (the returned index value) of less than `1` indicates that the community tends to recover faster from perturbations, assuming the interspecific interaction strengths and self-regulation effects are constant. Therefore, `1` represents a threshold above which the system is unstable and below which the system is stable. However, any persistent increase in Jacobian index is of concern for resilience loss. 

<br>  

### _Usage - Univariate_
The univariate form of Jacobian estimated stability is performed using the `uniJI()` function. Following Grziwotz *et al.* (2023), `uniJI()` estimates the strength of time embedded relationship between a univariate time series and lagged forms of itself.

In `uniJI()`, we again provide community data but for a single species (first column being time, focal species the second). We must also provide a `winsize`, embedding dimension (`E`) and `tau`. `E` and `tau` are complicated to parameterise and requires a degree of exploration. Initially, we suggest `E` to be in the range 1 to 6, and `tau` in the range 3 - 12, but refer readers to the excellent introduction by Chang *et al.* (2017) for further details.

```{r uniJI,message=FALSE, warning=FALSE,results='hide'}
eg_uniJI <- uniJI(data = simTransComms$community1[,2:3], winsize = 25, E = 3)
```

Plotting the index for this species, we observe a loss of stability prior to the transition and dramatic overcompensation following it.

```{r uniJI2,dpi=144}
uniJI_plot_data <- merge(simTransComms$community1,eg_uniJI,by="time") # combine the mvi data with the raw time series

ggplot(uniJI_plot_data,aes(x = time, y = smap_J)) + geom_line() + geom_vline(xintercept = uniJI_plot_data$inflection_pt,col = "grey60", linetype = "dashed") + theme_bw() + xlab("Time") + ylab("Univariate stability index value")
```

Lets repeat this for all species to give us an overall idea of system resilience.

```{r uniJI_comp,message=FALSE, warning=FALSE,results='hide'}
all_spp_uniJI <- sapply(grep("spp_",colnames(simTransComms$community1)),FUN = function(x){
  if(x == 3){
  uniJI(data = simTransComms$community1[,c(2,x)], winsize = 25, E = 3)
  }else{
  uniJI(data = simTransComms$community1[,c(2,x)], winsize = 25, E = 3)[,2]
  }
  }) # for each species, calculate the univariate stability index

all_spp_uniJI <- do.call("cbind",all_spp_uniJI) # merge the list in to a data.frame
names(all_spp_uniJI) <- c("time",paste("spp",1:5,sep="_")) # and rename missing columns

all_spp_plot_data <- merge(tidyr::pivot_longer(all_spp_uniJI,cols = -time,names_to = "species",values_to = "smap_J"), simTransComms$community1[,c("time","inflection_pt")], by = "time") #pivot_longer for easier plotting and merge with the inflection point data

ggplot(all_spp_plot_data,aes(x = time, y = smap_J)) + geom_line(aes(col=species)) + geom_vline(xintercept = all_spp_plot_data$inflection_pt,col = "grey60", linetype = "dashed") + theme_bw() + xlab("Time") + ylab("Univariate stability index value") + scale_colour_manual(values = c("black","#6886c4","#bfbd3d","#69c756","#e281fe")) + facet_wrap(~species)
```

This approach highlights how each species responds uniquely and can help identify the most sensitive species. I.e which are possible indicator species.

<br>  

### _Usage - Multivariate_
Conversely, to compute the multivariate form of Jacobian estimation, we can use the function `multiJI()`. `multiJI()` fits a local linear model between all species which predicts the system's future value from the reconstructed state-space. The resulting coefficients of this model are therefore a proxy for the interaction strength between species. 

Note `E` and `tau` are not required for the `EWSmethods` form of index. For `E`, the number of species themselves are used for embedding, though this does limit the minimum length of time series to being the number of species. Instead, `theta` is impactful, and represents the tuning parameter for forecasts (if not provided, `multiJI()` identifies the optimal `theta` using root mean square error). Conversely, `tau` is held constant at `-1` as there is no self embedding due to the index being a multivariate measure. 

> ⚠️ **Warning!** - 
> *`multiJI()` is significantly slower than any of the other resilience indicators provided by `EWSMethods` due to the multiple cross referencing required between species. Therefore increasing the number of species and length of time series (or decreasing `winsize`) can dramatically increase computation time*.

<br>  

```{r multiJI,message=FALSE, warning=FALSE,results='hide'}
eg_multiJI <- multiJI(data = simTransComms$community1[,2:7], winsize = 25)
```

Plotting the index for this species, we observe a loss of stability significantly prior to the transition and much earlier than the `uniJI()` estimated index.

```{r multiJI2,dpi=144}
multiJI_plot_data <- merge(simTransComms$community1,eg_multiJI,by="time") # combine the mvi data with the raw time series

ggplot(multiJI_plot_data,aes(x = time, y = smap_J)) + geom_line() + geom_vline(xintercept = multiJI_plot_data$inflection_pt,col = "grey60", linetype = "dashed") + theme_bw() + xlab("Time") + ylab("Mulivariate stability index value")
```

<br>  

### *Caveats* 
Both s-map methods are sensitive to the choice of `E` and `tau`. Additionally, in seasonal/cyclical data, the shared cycles or seasons across species can artificially inflate estimates of causality and relatedness. In these circumstances, it is recommended to detrend data prior to estimating these resilience metrics (Ushio *et al.* 2018). `EWSmethods` provides the capability to detrend and deaseason data via the functions `detrend_ts()` and `deseason_ts()` respectively. Similarly, it is sensible to scale/normalise all time series to be of equivalent magnitude to aid model fitting. A `scale = TRUE` argument is present in both `uniJI()` and `multiJI()` to achieve this.

Additionally, the two approaches require time series with no extended period of unchanging values (specifically a period equal to or longer than the window size). If `uniJI()` or `multiJI()` detect such a circumstance, the current iteration returns `NA` and the time series is removed respectively. For `multiJI()` if there are more species than data points in the window, then the index will fail. This is due to a multivariate index calculated across `n` species time delays up to lag `n-1`, which may not be possible if the time series are short. In this circumstance, we recommend sampling a subset of species, either randomly or the most abundance species (those less likely to have a period of unchanging values).

<br>  

# 3) Fisher Information

Fisher Information (FI) attempts to estimate the amount of information data can provide on an unmeasured parameter (Fisher and Russell 1922). It consequently has been applied as an indicator of system dynamics and stability (Ahmad et al. 2016). `EWSmethods` provides a simplified discrete time form of Fisher and Russel (1922)’s mathematic proof following Karunanithi *et al.* (2008):

\begin{equation}
FI\approx 4 \sum_{i=1}^{m}[q_i-q_{(i+1)}]^2
\end{equation}

where $q_{i}^2$ is the amplitude of the probability of observing states of the system at time
window $i$ and $m$ is the number of possible ‘states’. Possible states are defining by comparing the difference between temporally adjacent data windows to a reference ‘uncertainty’. If the absolute difference in density is less than the reference deviation for all variables (i.e. species densities), then the windows are binned in to the same ‘state’.

<br>  

### *Interpretation* 
Decreasing FI represents decreasing stability.

<br>  

### *Usage*
Provide the community data (first column being time, the remainder being species values) to the `FI()` function, specify a `winsize` to calculate the index over, and a size-of-states vector (`sost`). The default `winsize` is 50\% the total length of the time series, and `sost` is typically suggested to be the standard deviation of each species across the entire time series multiplied by 2 (Karunanithi et al. 2008). `winspace` indicates the number of data points to slide the rolling window along (`1` is standard) and `TL` is the 'tightening level'. `TL` is the percentage of points shared between states and allows the algorithm to classify data points to/from the same state.

```{r fi}
eg.sost <- t(apply(simTransComms$community1[,3:7], MARGIN = 2, FUN = sd)) # define size-of-states using the standard deviation of each species separately. Must be wide format hence t()

egFI <- FI(data = simTransComms$community1[,2:7], sost =  eg.sost, winsize = 25,winspace = 1, TL = 90)$FI
```

If we again plot the index, we can see that before and after the 'inflection point' (where the community goes through a transition) there are clear differences in stability. 

What is also apparent, is that the FI begins to decrease prior to the sudden change, and stability only returns after an extended period post constant stress (stress is constant from *t200* onward).

```{r fi2,dpi=144}
fi_plot_data <- merge(simTransComms$community1,egFI,by="time") # combine the mvi data with the raw time series

ggplot(fi_plot_data,aes(x = time, y = FI)) + geom_line() + geom_vline(xintercept = fi_plot_data$inflection_pt,col = "grey60", linetype = "dashed") + theme_bw() + xlab("Time") + ylab("Fisher Information value")
```

<br>  

### *Caveats*
FI is very sensitive to the size-of-states and tightening level. There is also no threshold of stable vs unstable as it is the relative change of FI that is important, not the absolute value.

<br>  

# References

Ahmad N., Derrible S., Eason T., and Cabezas H. (2016) Using Fisher information to track stability in multivariate systems. *Royal Society Open Science*. 3, 11, 160582. [doi: 10.1098/rsos.160582](https://royalsocietypublishing.org/doi/10.1098/rsos.160582)

Brock, W., & Carpenter, S. (2006). Variance as a leading indicator of regime shift in ecosystem services. *Ecology and Society*, 11, 9. [http://www.ecologyandsociety.org/vol11/iss2/art9/](https://www.ecologyandsociety.org/vol11/iss2/art9/)

Chang, C.-W., Ushio, M., and Hsieh, C.-h. (2017). Empirical Dynamic Modeling for beginners. *Ecological Research*, 32,6,785–796. [doi:10.1007/s11284-017-1469-9](https://esj-journals.onlinelibrary.wiley.com/doi/full/10.1007/s11284-017-1469-9)

Fisher, R. A., & Russell, E. J. (1922). On the mathematical foundations of theoretical statistics. *Philosophical Transactions of the Royal Society of London Series A: Containing Papers of a Mathematical or Physical Character*, 222, 309–368. [doi:10.1098/rsta.1922.0009](https://royalsocietypublishing.org/doi/10.1098/rsta.1922.0009)

Grziwotz, F., Chang, C.-W., Dakos, V., et al. (2023). Anticipating the occurrence and type of critical transitions. *Science Advances*, 9. [doi:10.1126/sciadv.abq4558](https://www.science.org/doi/10.1126/sciadv.abq4558)

Karunanithi, A. T., Cabezas, H., Frieden, et al. (2008). Detection and assessment of ecosystem regime shifts from Fisher Information. *Ecology and Society*, 13, 1. [https://www.jstor.org/stable/26267924](https://www.jstor.org/stable/26267924)

Sugihara, G., May, R., Ye, H., Hsieh, C., Deyle, E., Fogarty, M., & Munch, S. (2012). Detecting causality in complex ecosystems. *Science*, 338, 496–500. [doi:10.1126/science.1227079](https://www.science.org/doi/10.1126/science.1227079)

Takens, F. (1981). Detecting strange attractors in turbulence. *Dynamical Systems and Turbulence*, Warwick 1980. Ed. by D. Rand and L.-S. Young. Berlin, Heidelberg: Springer Berlin Heidelberg, pp. 366–381 

Ushio, M., Hsieh, Ch., Masuda, R. et al. (2018) Fluctuating interaction network and time-varying stability of a natural fish community. *Nature*, 554, 360–363. [doi:10.1038/nature25504](https://www.nature.com/articles/nature25504)
